---
title: "Untitled"
author: "Lionel Forbes"
date: "2/19/2020"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
str(surveys)
view(surveys)
```

```{r}
select(surveys, plot_id, species_id, weight)
```
```{r}
select(surveys, -record_id, -species_id)
```

```{r}
filter(surveys, year == 1995)
```

```{r}
surveys <- filter(surveys,weight < 5)
surveys_sml <- select(surveys, species_id, sex, weight)
```

```{r}
surveys_sml <- surveys %>% 
  filter(weight < 5) %>% 
  select(species_id, sex, weight)

surveys_sml
```

```{r}
surveys %>% 
  select(year,sex,weight) %>% 
  filter(year == 1985)
```

```{r}
surveys %>% 
  mutate(weight_kg = weight / 1000)
```

```{r}
surveys %>% 
  mutate(weight_kg = weight / 1000,
         weight_lb = weight_kg * 2.2) %>% 
head()
```

```{r}
surveys %>%
  filter(!is.na(weight)) %>%
  mutate(weight_kg = weight / 1000) %>%
  head()
```

Challenge 2
```{r}
surveys_hindfoot_half <- surveys %>%
    filter(!is.na(hindfoot_length)) %>%
    mutate(hindfoot_half = hindfoot_length / 2) %>%
  filter(hindfoot_half < 30) %>% 
  select(species_id, hindfoot_half)

surveys_hindfoot_half
```

```{r}
surveys %>%
  group_by(sex) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE))

```

```{r}
surveys %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE))
```

surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight))

surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight)) %>%
  print(n = 15)
  
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            min_weight = min(weight))

surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            min_weight = min(weight)) %>%
  arrange(min_weight)
  
  surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            min_weight = min(weight)) %>%
  arrange(desc(mean_weight))
  
surveys %>%
    count(sex)

surveys %>%
    group_by(sex) %>%
    summarise(count = n())
    
surveys %>%
    count(sex, sort = FALSE) 
    
surveys %>%
  count(sex, species) %>%
  arrange(species, desc(n))

Challenge 3
```{r}
surveys %>% 
  count(plot_type)

surveys %>%
    filter(!is.na(hindfoot_length)) %>%
    group_by(species_id) %>%
    summarize(
        mean_hindfoot_length = mean(hindfoot_length),
        min_hindfoot_length = min(hindfoot_length),
        max_hindfoot_length = max(hindfoot_length),
        n = n()
    )

surveys %>%
    filter(!is.na(weight)) %>%
    group_by(year) %>%
    filter(weight == max(weight)) %>%
    select(year, genus, species, weight) %>%
    arrange(year)
```

```{r}
surveys_gw <- surveys %>%
  filter(!is.na(weight)) %>%
  group_by(genus, plot_id) %>%
  summarize(mean_weight = mean(weight))

str(surveys_gw)
```


```{r}
surveys_spread <- surveys_gw %>%
  spread(key = genus, value = mean_weight)

str(surveys_spread)

surveys_gw %>%
  spread(genus, mean_weight, fill = 0) %>%
  head()

```

```{r}
surveys_gather <- surveys_spread %>%
  gather(key = genus, value = mean_weight, -plot_id)

str(surveys_gather)
```


Challenge 4
```{r}
rich_time <- surveys %>%
  group_by(plot_id, year) %>%
  summarize(n_genera = n_distinct(genus)) %>%
  spread(year, n_genera)

head(rich_time)
```

```{r}
rich_time %>%
  gather(year, n_genera, -plot_id)
```

```{r}
surveys_long <- surveys %>%
  gather(measurement, value, hindfoot_length, weight)
```

```{r}
surveys_long %>%
  group_by(year, measurement, plot_type) %>%
  summarize(mean_value = mean(value, na.rm=TRUE)) %>%
  spread(measurement, mean_value)
```


Exercise 1
```{r}
SNPs <- read_tsv("data/23andMe_complete.txt", skip = 14, col_types = 
  cols(chromosome = col_factor()))

Complete <- select(SNPs, position, genotype)

Complete

```

Exercise 2
```{r}
SNPs %>% 
  filter(!genotype %in% c("A", "T", "--")) %>% 
  filter(chromosome != "MT")

```

Exercise 3
```{r}
SNPs %>% 
  bgroup_by(position, chromosome) %>% 
  summarise(min_position = min(position), max_position = max(position)) 
```
```{r}
SNPs %>%
  count(position)
```

Exercise 4


```{r}
SNPs %>% 
  summarise(min_position = min(position), max_position = max(position)) %>% 
  mutate(density = max_position - min_position )
```


